---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## T-Cell signalling dataset

The T-cell signalling dataset consists of simultaneous measurements of 11 phosphorylated proteins and phospholypids derived from thousands of individual primary immune system cells, specifically T cells. When T cells are stimulated, the signal flows across a series of physical interactions between the measured proteins. The network of these interactions forms the T cell signalling pathway.  

A directed acyclic network can be used to represent this signalling pathway.  The authors of this publication shared constructed a reference network, which I stored as an R object in my package [bnlearn](#): 

```{r}
library(dplyr)
coefs <- centered_scaled_gauss %>% coef 
parents <- melt(lapply(coef(centered_scaled_gauss), names))$value
coefs %>%
  melt %>%
  mutate(child = L1, weight = value) %>%
  mutate(parent = parents) %>%
  select(parent, child, weight)
```

tcell_examples


The loaded R object contains both the raw data and preprocessed data used in the original publication for network inference.  

The data cotains information on interventions that provide signal to the network as well as activate and inhibit individual proteins.  These interventions enable network inference to provide edges with a causal interpretation. Run *?data* for more background on the data.

```{r}
raw <- tcells$raw
processed <- tcells$processed
```

The data frame *processed* has intervention information in the INT column.

## Network inference experiment workflow

The workflow for network inference experiments in this context is as follows:

1 Select proteins
2 Quantify data
3 Apply network inference with model averaging
4 Evaluate strength of edges
5 Validate both highly ranked and interesting edges in follow up experiments

Mario Scutari describes an R-based workflow on this T-cell data.  Using *bnlearn* 

```{r}
scutari_analysis <- function(df){
  int_array <- as.numeric(df$INT); df$INT <- NULL  # Pull out the intervetions, so only the proteins remain.
  int_arg <- lapply(seq_along(df), function(i){
   which(int_array == i)}) %>%
   structure(names =  names(df))
   random.graph(nodes = names(df), # Generate random graph
                               method = "melancon",
                               num = 500,
                               burn.in = 10^5,
                               every = 100) %>%
    lapply(function(net){ # Fit Tabu search to each graph
      tabu(df, score = "mbde", exp = int_arg, iss = 10, start = net, tabu = 50)
    })  %>%
    custom.strength(nodes = names(df)) # Aggregate averaging results
}
model_averaging_results <- scutari_analysis(processed)

```

## Description of experiment

Assume Sachs is ground truth
My theory tells me that when the input network has spurious edges, my model will perform well by giving weak weights. Using a Gaussian model, or model averaging, some of Sach's edges will be weak.  The edges exist, but 
signal flow is low.  I want those weak edges in the network so I can shrink them.  
Suppose I remove Mek and Raf i.e. the Mapk pathway.  Then the influence on Erk then comes only PKA.  
I am supposed to be quantifying different proteins.
We pick different sets of nodes, and for each edge plot a roc curve

## Network inference workflow

Scutari book details an effective workflow for network inference on this dataset using the bnlearn package.  





