---
title: "Converting KEGG model to a Signal Graph Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r}
library(signalgraph)
```

Load the processed T-cell data.

```{r}
data(tcells, package = "bninfo")
```

The T-cell data uses protein names.  I need to map these to KEGG names

```{r}
protein_names <- names(tcells$raw_data$cd3cd28)
sachs_kegg_map <- data.frame(
  label = c("Raf", "Mek", "Plcg", "Erk", "Akt", "PKC", "P38", "Jnk", "PKA"),
  kegg = c("hsa:5894", "hsa:5604", "hsa:5335", "hsa:5594", "hsa:10000", "hsa:5588", "hsa:1432", 
           "hsa:5601", "hsa:5566"),
  stringsAsFactors = F
  ) %>%
  `rownames<-`(.$kegg)
sachs_kegg_map
```

I use KEGGgraph to grab the T cell receptor signaling pathway.  By default this stores nodes by KEGG IDs.  With some tweaking I can grab the gene names for a more convenient naming scheme.  The downloaded graph is a graphNEL, I make the conversion to igraph.


```{r}
library(KEGGgraph)
library(dplyr)
maps <- c(tcell = "04660", mapk = "04010", P13k = "04151", ras = "04014")#, rap1 = "04015")
vertex_master <- NULL
edge_master <- NULL
for(map in maps){
  g_nell <- tempfile() %T>%
  {retrieveKGML(map, organism="hsa", destfile=., method="curl", quiet=TRUE)} %>%
  parseKGML2Graph(expandGenes=FALSE) 
  vertex_list <- getKEGGnodeData(g_nell) %>%
    {data.frame(
      kegg = unlist(lapply(., function(item) item@name[1])),
      label = gsub("\\.", "", unlist(
        lapply(., function(item) strsplit(item@graphics@name, ",")[[1]][1]))),
      stringsAsFactors = F)}
  g <- igraph.from.graphNEL(g_nell) 
  V(g)$name <- vertex_list$kegg 
  edge_list <- getKEGGedgeData(g_nell) %>%
    lapply(function(item){
      if(length(item@subtype) > 0) {
        subtype_info <- item@subtype
        # KEGG uses a hierarchy of term for describing terms
        # for example, the first edge type is "activation", the second is "phosphorylation"
        # where phosphorylation is a type of activation.  The second term is more specific than
        # the first, so when it is provided, use it in lieu of the first type.
        if(length(subtype_info) > 1) {
          return(subtype_info[[2]]@name)
        } 
        return(subtype_info$subtype@name)
      }
      NA
    }) %>%
    unlist %>%
    {cbind(get.edgelist(g), type = .)} %>%
    data.frame
  edge_master <- rbind(edge_master, edge_list)
  vertex_master <- rbind(vertex_master, vertex_list)
}
phospho_edges <- edge_master %>%
  as.data.frame %>%
  {.$V1[.$V1 == "hsa:5601"] <-  "hsa:5599"; .} %>% # 5601 and 5999 are both JNK.  Use only one JNK
  {.$V2[.$V2 == "hsa:5601"] <- "hsa:5599"; .} %>%
  {.$V1[.$V1 %in% c("hsa:5578", "hsa:5579", "hsa:5582")] <- "hsa:5588"; .} %>%
  {.$V2[.$V2 %in% c("hsa:5578", "hsa:5579", "hsa:5582")] <- "hsa:5588"; .} %>%
  unique %>%
  filter(type == "phosphorylation")
  
  #filter(!(type %in% c("indirect effect", "dissociation", "compound", "dephosphorylation", "binding/association")))  
vertex_master <- vertex_master %>%
  as.data.frame %>%
  {.$kegg[.$kegg == "hsa:5601"] <- "hsa:5599"; .} %>%
  {.$label[.$kegg == "hsa:5599"] <- "Jnk"; .} %>%
  # Merge the PKC's
  {.$kegg[.$kegg %in% c("hsa:5578", "hsa:5579", "hsa:5582")] <- "hsa:5588"; .} %>%
  {.$label[.$kegg == "hsa:5588"] <- "PKC"; .} %>%
  unique 

g_igraph <- graph.data.frame(phospho_edges, directed = TRUE, vertices = vertex_master) %>%
  simplify(edge.attr.comb = "concat")
```

The gene names are informative but not unique, the KEGG id's vice versa, so I combine the two.  The proteins in the data keep their original names.  The trisphosphate compounds PIP2 and PIP3 are in the data, but in KEGG they are a [compound](http://www.kegg.jp/dbget-bin/www_bget?C01245).  I add these back in, as well as protein kinase A (PKA) and some other interactions from other pathways (see references below).  Then I add in the interventions in the data to the graph.

```{r}
# Give use labels instead of kegg ids
V(g_igraph)$kegg_id <- V(g_igraph)$name
V(g_igraph)$name <- V(g_igraph)$label
# Rename ones in data
v_proteins <- V(g_igraph)[kegg_id %in% sachs_kegg_map$kegg] 
V(g_igraph)[v_proteins]$name <- sachs_kegg_map[v_proteins$kegg_id, "label"]
# Change the name of other key pathway proteins
#V(g_igraph)[keepers[!is.na(keepers$label)]]$name <- V(g_igraph)[keepers[!is.na(keepers$label)]]$label

# "hsa:23533" is "PIK3R5" is PI3K

# Add in the missing vertices and edges
V(g_igraph)["PIK3R5"]$name <- "PI3K"
g_igraph <- g_igraph +
  vertices(c("PIP2", "PIP3")) + # add in PI3K and phosphatidylinositols
  igraph::edges(
    c("PI3K", "PIP3",
      "PI3K", "PIP2",
      "Plcg", "PIP3", # add in phosphatidylinositols edges
      "PIP3", "PIP2", 
      "Plcg", "PIP2",
      "PIP3", "Akt")) 

# + 
#   igraph::edges(
#     c("PKC", "MAP3K1",
#       "PKC", "MAP2K4", 
#       "PKA", "MAP3K1"))

#V(g_igraph)["hsa:5584"]$name <- "PKCI"
#Edges not in KEGG suggested by Sachs Paper
# c("PKC", "MP")
# new_edges <- c(V(g_igraph)["PKC"], V(g_igraph)["MAP3K1"], 
#            V(g_igraph)["PKC"], V(g_igraph)["hsa:6416"],
#            V(g_igraph)["PKA"], V(g_igraph)["hsa:4214"], 
#            V(g_igraph)["PIP3"], V(g_igraph)["Akt"], 
#            V(g_igraph)["hsa:4214"], V(g_igraph)["hsa:5599"],
#            V(g_igraph)["hsa:6416"], V(g_igraph)["Jnk"], 
#            V(g_igraph)["PKC"], V(g_igraph)["hsa:22800"], 
#            V(g_igraph)["hsa:22800"],V(g_igraph)["Raf"])
# g_old <- g_igraph
# g_igraph <- add_edges(g_igraph, new_edges)
# g_igraph <- add_vertices(g_igraph, 1, attr = list(name = "Raf497", label = "Raf497"))
# g_igraph <- add_edges(g_igraph, c(V(g_igraph)["PKC"], V(g_igraph)["Raf497"],
#                                   V(g_igraph)["Raf497"], V(g_igraph)["Mek"]))
```

I remove the unlinked vertices.

```{r}
g_igraph <- g_igraph - V(g_igraph)[igraph::degree(g_igraph) == 0]
```


Plotting with RGraphviz.  The green nodes are are the proteins observed in the data.

```{r}
node_list <- rep("green", length(protein_names)) %>% 
  structure(names = V(g_igraph)[protein_names]$name) %>%
  {list(fill = ., fontsize=0.01)} 
g_igraph %>% # Give vertices names if they do not have nay 
  igraph.to.graphNEL(.) %>% # convert to a graphNEL
  {Rgraphviz::layoutGraph(.)} %>% # lay the graph out
  {graph::`nodeRenderInfo<-`(., node_list)} %>%
  {Rgraphviz::renderGraph(.)} # Render the graph
```

```{r}
V(g_igraph)["RRAS2"]$name <- "Ras"

g_igraph <- g_igraph + igraph::edges(
  c("Ras", "Raf",
    "PKC", "Ras", # This  is a phosphorylation, not sure how was missed
    "PKC", "MAP3K7", # A non-phospho edge, adding it preserves PKC link to JNK implied by Sachs via Map3k7 and map2k7
    "MAP3K7", "MAP2K7", # This should have been recovered when I combined JNKs.  Not sure of error.
    "PKA", "MAP3K1",
    "PKA", "Raf", # Preserve PKA link to Raf via non-phospho edge with Rap1
    "PKC", "MAP3K5"))# No evidence of this in KEGG, but Sachs paper and Sachs data suggests it
  # or perhaps leave this out and suggest the method "discovered" it in the data
  # Inferred between PKC and PKA in paper is justified via a PKC ->cAMP -> PKA relationship.  Cant
  # find evidence of this in KEGG.  Leaving it out.
  # Lots of other edges
g_igraph <- tryCatch(delete_edges(g_igraph,  E(g_igraph)[from("MAP2K7")][to("MAP3K7")] ),
                     error = function(e) g_igraph) # Not sure how this got in there.
  
  
# add a second Raf directly to the model


#            V(g_igraph)["hsa:4214"], V(g_igraph)["hsa:5599"],
#            V(g_igraph)["hsa:6416"], V(g_igraph)["Jnk"], 
#            V(g_igraph)["PKC"], V(g_igraph)["hsa:22800"], 
#            V(g_igraph)["hsa:22800"],V(g_igraph)["Raf"])                                 
                                 
# LFA-1 intervention on RAS
direct_interventions <- c("act_RAS", "act_PKA", "act_PI3K", "act_PKC", "ihb_PIP2")
g_igraph <- g_igraph + 
  vertices(direct_interventions) + 
  igraph::edges(
    c("act_RAS", "Ras",
      "act_PKA", "PKA",
      "act_PI3K", "PI3K",
      "act_PKC", "PKC",
      "ihb_PIP2", "PIP2"))
node_list <- c(rep("green", length(protein_names)),
               rep("orange", length(direct_interventions))) %>% 
  structure(names = c(protein_names, direct_interventions)) %>%
  {list(fill = ., fontsize = .01)} 
g_igraph %>% # Give vertices names if they do not have nay 
  igraph.to.graphNEL %>% # convert to a graphNEL
  {Rgraphviz::layoutGraph(.)} %>% # lay the graph out
  {graph::`nodeRenderInfo<-`(., node_list)} %>%
  {Rgraphviz::renderGraph(.)} # Render the graph
```

To map how signal flows between these highlighted nodes, I zoom in on nodes that are both downstream and upstream of nodes that have data.


```{r}
observed <- V(g_igraph)[c(protein_names, direct_interventions)]
unobserved <- V(g_igraph)[setdiff(V(g_igraph)$name, observed$name)]
keepers <- unobserved[sapply(unobserved, function(B){
  is_downstream <- any(sapply(observed, function(A) isBDownstreamOfA(g_igraph, A, B)))
  is_upstream <- any(sapply(observed, function(A) isBUpstreamOfA(g_igraph, A, B)))
  is_downstream && is_upstream
})]
V(g_igraph)[keepers[!is.na(keepers$label)]]$name <- V(g_igraph)[keepers[!is.na(keepers$label)]]$label
tossers <- V(g_igraph)[setdiff(V(g_igraph), c(observed, keepers))]
sub_net <- delete_vertices(g_igraph, tossers)
V(sub_net)["RRAS2"]$name <- "Ras"
V(sub_net)["PIK3R5"]$name <- "PI3K"
node_list <- c(rep("green", length(protein_names)),
               rep("orange", length(direct_interventions))) %>% 
  structure(names = c(protein_names, direct_interventions)) %>%
  {list(fill = ., fontsize = 30)} 
sub_net %>%
  igraph.to.graphNEL %>%
  layoutGraph %>%
  {graph::`nodeRenderInfo<-`(., node_list)} %>%
  renderGraph
```

```{r}
activity_blockers <- paste0("ihb_", c("Mek", "Akt", "PKC", "PI3K"))
activity_blocker_edges <- lapply(V(sub_net)[c("Mek", "Akt", "PKC", "PI3K")], function(v){
  inhibition <- paste0("ihb_", V(sub_net)[v]$name)
  cbind(inhibition, V(sub_net)[ichildren(sub_net, v)]$name)
  }) %>% 
  {do.call("rbind", .)} %>%
  t %>%
  as.character
sub_net <- sub_net + 
  vertices(activity_blockers) +
  igraph::edges(activity_blocker_edges)
ints <- c(direct_interventions, activity_blockers)
node_list <- c(rep("green", length(protein_names)),
               rep("orange", length(ints))) %>% 
  structure(names = c(protein_names, ints)) %>%
  {list(fill = ., fontsize = 30)} 
sub_net %>%
  igraph.to.graphNEL %>%
  layoutGraph %>%
  {graph::`nodeRenderInfo<-`(., node_list)} %>%
  renderGraph  
```



```{r}
#icam2 = act_RAS, aktinhib = ihb_Akt, g0076 = ihb_PKC, psit = ihb_PIP2, U0126 = ihb_Mek,
#ly = ihb_PI3k, pma = act_PKC, b2camp = act_PKA, cd3cd28 = act_PI3K
# Might have to add iterventions to children
devtools::load_all()
.data_list <- tcells$raw_data %>%
  lapply(function(.data){
    log(.data) %>% 
      rescale_df %$%
      df %>%
      mutate(act_RAS = 0, act_PKA = 0, act_PKC = 0, act_PI3K = 1, ihb_Akt = 0, ihb_PKC = 0, 
           ihb_PIP2 = 0, ihb_Mek = 0, ihb_PI3K = 0)
    })
icam_set <- c("cd3cd28icam2+aktinhib", "cd3cd28icam2+g0076", "cd3cd28icam2+psit", 
             "cd3cd28icam2+u0126", "cd3cd28icam2+ly", "cd3cd28icam2")
.data_list[icam_set] <- .data_list[icam_set] %>%
  lapply(function(.data){
      mutate(.data, act_RAS = 1, ihb_Akt = 0, ihb_PKC = 0, 
           ihb_PIP2 = 0, ihb_Mek = 0, act_PKA = 0)
  })
# Remove RAS and PI3K activation from PKA and PKC activation datasets. Add PKA and PKC activation
.data_list[c("pma", "b2camp" )] <- .data_list[c("pma", "b2camp" )] %>%
  lapply(function(.data){
    mutate(.data, act_PKC = 1, act_PKA = 0, act_RAS = 0, act_PI3K = 0)
  })
.data_list$b2camp$act_PKC <- 0
.data_list$b2camp$act_PKA <- 1

# Add inhibitions
.data_list[["cd3cd28icam2+aktinhib"]]$ihb_Akt <- 1
.data_list[["cd3cd28+aktinhib"]]$ihb_Akt <- 1
.data_list[["cd3cd28icam2+g0076"]]$ihb_PKC <- 1
.data_list[["cd3cd28+g0076"]]$ihb_PKC <- 1
.data_list[["cd3cd28icam2+psit"]]$ihb_PIP2 <- 1
.data_list[["cd3cd28+psitect"]]$ihb_PIP2 <- 1
.data_list[["cd3cd28icam2+u0126"]]$ihb_Mek <- 1
.data_list[["cd3cd28+u0126"]]$ihb_Mek <- 1
.data_list[["cd3cd28icam2+ly"]]$ihb_PI3K <- 1
.data_list[["cd3cd28+ly"]]$ihb_PI3K <- 1
.data <- as.data.frame(do.call("rbind", .data_list))
```

```{r}
sub_net <- simplify(sub_net)
.sub_data <- sample_n(.data, 500)
system.time(fit1 <- fitNetwork(sub_net, .sub_data, fixed = ints, graph_attr = list(L2_pen = .01,
                                                               min.max.constraints = c(-50, 50)), 
                   max.iter = 1, verbose = TRUE))
getMSE(fit1) # 
vertexMSEs(fit1)
sg_viz(fit1)
qqnorm(unlist(V(fit1)["Erk"]$observed) - unlist(V(fit1)["Erk"]$output.signal))
  fit11 <- fit_initialized_sg(fit1, max.iter = 1, verbose = TRUE)
getMSE(fit11)
sg_viz(fit11)
fit111 <- fit_initialized_sg(fit11, min.iter = 3, verbose = TRUE)
# Also check errors
fit2 <- fitNetwork(sub_net, .sub_data, fixed = ints, graph_attr = list(L2_pen = .01,
                                                               min.max.constraints = c(-50, 50)), 
                   max.iter = 1, verbose = TRUE)
fit21 <- fit_initialized_sg(fit2, max.iter = 1, verbose = TRUE)
fit3 <- fitNetwork(sub_net, .sub_data, fixed = ints, graph_attr = list(L1_pen = .01, L2_pen = .01,
                                                               min.max.constraints = c(-50, 50)), 
                   max.iter = 1, verbose = TRUE)
fit31 <- fit_initialized_sg(fit2, max.iter = 1, verbose = TRUE)
fit4 <- fitNetwork(sub_net, .sub_data, fixed = ints, graph_attr = list(L2_pen = .005,
                                                               min.max.constraints = c(-50, 50)), max.iter = 1, verbose = TRUE)
fit5 <- fitNetwork(sub_net, .sub_data, fixed = ints, graph_attr = list(L1_pen = .005, L2_pen = .005,
                                                               min.max.constraints = c(-50, 50)), max.iter = 1, verbose = TRUE)
```


### References
Sachs, Karen, et al. "Causal protein-signaling networks derived from multiparameter single-cell data." Science 308.5721 (2005): 523-529. [dataset download](http://www.sciencemag.org/content/suppl/2005/04/21/308.5721.523.DC1/Sachs.SOM.Datasets.zip)

Marsland, B. J. & Kopf, M. T-cell fate and function: PKC-theta and beyond. Trends Immunol. 29, 179-185 (2008)

Sakaguchi, Shimon. "Naturally arising CD4+ regulatory T cells for immunologic self-tolerance and negative control of immune responses." Annu. Rev. Immunol. 22 (2004): 531-562.

Perez, Omar D., et al. "Leukocyte functional antigen 1 lowers T cell activation thresholds and signaling through cytohesin-1 and Jun-activating binding protein 1." Nature immunology 4.11 (2003): 1083-1092.

Altan-Bonnet, Grégoire, and Ronald N. Germain. "Modeling T cell antigen discrimination based on feedback control of digital ERK responses." PLoS biology 3.11 (2005): e356.

Das, Jayajit, et al. "Digital signaling and hysteresis characterize ras activation in lymphoid cells." Cell 136.2 (2009): 337-351.

Lipniacki, Tomasz, et al. "Stochastic effects and bistability in T cell receptor signaling." Journal of theoretical Biology 254.1 (2008): 110-122.

Nika, Konstantina, et al. "Haematopoietic protein tyrosine phosphatase (HePTP) phosphorylation by cAMP-dependent protein kinase in T-cells: dynamics and subcellular location." Biochem. J 378 (2004): 335-342.

Kolch, Walter, Muffy Calder, and David Gilbert. "When kinases meet mathematics: the systems biology of MAPK signalling." FEBS letters 579.8 (2005): 1891-1895.

Whitehurst, Angelique, Melanie H. Cobb, and Michael A. White. "Stimulus-coupled spatial restriction of extracellular signal-regulated kinase 1/2 activity contributes to the specificity of signal-response pathways." Molecular and cellular biology 24.23 (2004): 10145-10150.

Ferrell Jr, James E. "Self-perpetuating states in signal transduction: positive feedback, double-negative feedback and bistability." Current opinion in cell biology 14.2 (2002): 140-148.

Li, Wei, et al. "Blocked signal transduction to the ERK and JNK protein kinases in anergic CD4+ T cells." Science 271.5253 (1996): 1272-1276.

Fields, Patrick E., Thomas F. Gajewski, and Frank W. Fitch. "Blocked Ras activation in anergic CD4+ T cells." Science 271.5253 (1996): 1276-1278.